---
title: "Survival analysis of patients with osteosarcoma"
author: "Léa ROGUE"
output:
  pdf_document:
    fig_caption: true
    latex_engine: xelatex
    toc: true
    toc_depth: '4'
    includes:
      before_body: tex_files/list_of_figures.tex
date: "2025-03-13"
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  tidy.opts = list(width.cutoff = 60), 
  tidy = TRUE, 
  fig.align = 'center'
)
```

\newpage

Intro

# Load librairies
```{r, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(survival)
library(forestplot)
library(survminer)
library(ggplot2)
library(ggsurvfit)
library(gridExtra)
library(tidyr)
library(stringr)
```

# Functions
```{r}
# Function to apply coxph model per columns
apply_coxph_model <- function(df) {
  # Empty df
  results <- data.frame()

  # Loop on each columns to apply coxph
  for (col in colnames(df)[3:ncol(df)]) {
    model <- coxph(Surv(OS.delay, OS.event) ~ df[[col]], data = df)
    
    # Extract results
    exp_coef <- summary(model)$coefficients[,"exp(coef)"]
    lower_ci <- summary(model)$conf.int[,"lower .95"]
    upper_ci <- summary(model)$conf.int[,"upper .95"]
    p_value <- summary(model)$coefficients[,"Pr(>|z|)"]
    results <- rbind(results, data.frame(
      Variable = col,
      HR = exp_coef,
      LowerCI = lower_ci,
      UpperCI = upper_ci,
      Pvalue = p_value
    ))
  }
  return(results)
}

# Function to create Kaplan-Meier plot
plot_km <- function(mod, df) {
  km_plot <- ggsurvplot(mod,
             data = df,      
             pval = TRUE,
             conf.int = FALSE,
             ggtheme = theme_bw(),
             palette = c("#E7B800", "#2E9FDF", "#FF6F61", "#4EBB92"))
    return(km_plot$plot)
}

# Function to generate forest plot from coxph analysis from the function apply_coxph_model
generate_forestplot <- function(data, Type) {
  if (Type == TRUE){
    # Order by Signature and p-value
    data <- data[order(data$Signature,  data$Pvalue),]
    data$Pvalue <- sprintf("%.4f", data$Pvalue)
    data %>%
      forestplot(
        labeltext = c(Signature, Variable, Pvalue),   
        mean = HR,
        lower = LowerCI,
        upper = UpperCI,
        grid = TRUE,
        zero = 1,
        col = fpColors(box = "black", line = "black"),
        hrzl_lines = TRUE,
        title = "Forest Plot for Cox Model",
        txt_gp = fpTxtGp(label = gpar(fontsize = 8)),
        ci.vertices = TRUE,
        boxsize = 0.1) %>%
      fp_set_zebra_style("#EFEFEF") %>% 
      fp_add_header(
        Signature = c("", "Cell type"),
        Variable = c("", "Genes"),
        Pvalue = c("", "p-value"))
  }
  else {
    # Order by p-value
    data <- data[order(data$Pvalue),]
    data$Pvalue <- sprintf("%.4f", data$Pvalue)
    data %>%
      forestplot(
        labeltext = c(Variable, Pvalue),   
        mean = HR,
        lower = LowerCI,
        upper = UpperCI,
        grid = TRUE,
        zero = 1,
        col = fpColors(box = "black", line = "black"),
        hrzl_lines = TRUE,
        title = "Forest Plot for Cox Model",
        txt_gp = fpTxtGp(label = gpar(fontsize = 8)),
        ci.vertices = TRUE,
        boxsize = 0.1) %>%
      fp_set_zebra_style("#EFEFEF") %>% 
      fp_add_header(
        Variable = c("", "Genes"),
        Pvalue = c("", "p-value"))    
  }
}
```

# Load and format data
```{r}
# Matrix with z-scores intensities and information on immune signature and CTA
df_expr_z_scores <- read.table("../results/expr_matrix_CTA_sign_imm_z_scores.tsv", sep = "\t", header = TRUE, check.names = FALSE)

# Matrix with average expression per cell types in z-scores
df_imm_z_scores <- read.table("../results/imm_sign_avg_z_scores.tsv", sep = "\t", header = TRUE, check.names = FALSE)
rownames(df_imm_z_scores) <- df_imm_z_scores$Signature
df_imm_z_scores <- as.data.frame(t(df_imm_z_scores[, -1]))
```


```{r}
# Create a better matrix
metadata <- read.table("../data/metadata.tsv", sep = "\t", header = F)
df_metadata <- as.data.frame(t(metadata))
colnames(df_metadata) <- df_metadata[1,]
df_metadata <- df_metadata[-1,]
colnames(df_metadata) <- gsub(" ", "_", colnames(df_metadata))
df_metadata$OS.event <- ifelse(grepl("Alive", df_metadata$status), 0, 1)
df_metadata$OS.delay <- as.numeric(str_extract(df_metadata$status, "\\d+"))
df_metadata_surv <- df_metadata %>% select(Patient, OS.delay, OS.event)

```

\newpage

# I. Immune cells survival analysis
This part concern the survival analysis with the expression of the immune cells. This is to observe the impact of each cell types on the survival probabilities of the patients.

## 1) All chondrosarcoma types
### a- Categorical data
Categorical data refer to HIGH or LOW expression based on the median of z-scores. If the score is > than median, HIGH is affected contrary to < than median, LOW is affected. In this section, we use Cox regression.
```{r, fig.keep='last', fig.cap="Forest plot for categorical immune cell types expression (n = 82)"}
# Transform to categorical data and take continuous data
# Computing medians to have categorical data
medians_sign <- apply(df_imm_z_scores, 2, median)

# Add categorical data, LOW is for under expressed genes (under the median) and HIGH upreg
df_cat_all <- df_imm_z_scores
for (col in colnames(df_cat_all)) {
  new_column_name <- paste(col, "cat", sep = " ")
  df_cat_all[[new_column_name]] <- ifelse(df_imm_z_scores[[col]] > medians_sign[col], "HIGH", "LOW")
}

# Select categories
df_cat_all <- df_cat_all[, grep("cat", colnames(df_cat_all))]

# Replace space by _
colnames(df_cat_all) <- gsub(" ", "_", colnames(df_cat_all))

# Add col for merging
df_cat_all$Patient <- rownames(df_cat_all)

# Merge df
df_survival_all <- merge(df_metadata_surv, df_cat_all, by = "Patient")
rownames(df_survival_all) <- df_survival_all$Patient
df_survival_all <- df_survival_all[, -1]

# Continuous data
df_cont_all <- df_imm_z_scores

# Replace space by _
colnames(df_cont_all) <- gsub(" ", "_", colnames(df_cont_all))

# Merge df
df_cont_all$Patient <- rownames(df_cont_all)
df_survival_cont <- merge(df_metadata_surv, df_cont_all, by = "Patient")
rownames(df_survival_cont) <- df_survival_cont$Patient
df_survival_cont <- df_survival_cont[, -1]

# CTA data
rownames(df_expr_z_scores) <- df_expr_z_scores$SYMBOL
df_expr_cta <- subset(df_expr_z_scores, CTA == "CTA")
df_expr_cta <- df_expr_cta[,!colnames(df_expr_cta) %in% c("Signature", "CTA", "SYMBOL")]
df_expr_cta <- as.data.frame(t(df_expr_cta))
df_expr_cta <- df_expr_cta[df_metadata_surv$Patient,]
df_expr_cta$Patient <- rownames(df_expr_cta)
df_expr_cta <- merge(df_metadata_surv, df_expr_cta, by = "Patient")
rownames(df_expr_cta) <- df_expr_cta$Patient
df_expr_cta <- df_expr_cta[, -1]

# Apply coxph model on the categorical variables for the immune signatures
results_cat <- apply_coxph_model(df_survival_all)
#write.table(results_cat, "../results/results_coxhp.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
generate_forestplot(data = results_cat, Type = FALSE)
```
We see that the p-value are high, maybe because of categorical data because Cox model rank the values.

### b- Continuous data
In this section, we use z-scores data with Cox model.
```{r, fig.cap="Forest plot for continuous immune cell types expression (n = 82)"}
# Apply coxph on continuous
results_cont <- apply_coxph_model(df_survival_cont)
#write.table(results_cont, "../results/results_coxph_var_cont.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
#pdf("../results/figures/forest_plots/all_indiv/forest_plot_all_patients_var_cont.pdf")
generate_forestplot(results_cont, Type = FALSE)
#dev.off()
```
The p-values are smaller than the model with categorical data so it's seems to be better with continuous data.

For all patients, we see that the hazard ratio (survival probabilities) are increased for Th2 cells, Immune checkpoints, aDC and Treg. So, more these cell types are present in the tumors, more the probability of death is higher, these cell types have a bad impact on survival. Contrary to eosinophils and DC, the HR decreased so more these cells are present in the tumors, the death probability decreased so they have a good impact on prognostic.

### c- Kaplan-Meier plots
This part compute Kaplan-Meier plot based on log rank test that use categorical data to see the trend of survival for the 2 groups (HIGH or LOW expression of the cell types)
This shows the graph for each significant HR ratio previously computed with Cox model.
```{r, fig.height=20, fig.width=15, fig.cap="Kaplan-Meier plots for immune cell types (n = 82)"}
#Create Kaplan-Meier plots with categorical data
p1 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ DC_cat, data = df_survival_all), df_survival_all)
p2 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Eosinophils_cat, data = df_survival_all), df_survival_all)
p3 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Immune_checkpoints_cat, data = df_survival_all), df_survival_all)
p4 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ TFH_cat, data = df_survival_all), df_survival_all)
p5 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Th2_cells_cat, data = df_survival_all), df_survival_all)
p6 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Treg_cat, data = df_survival_all), df_survival_all)
p7 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ aDC_cat, data = df_survival_all), df_survival_all)
#pdf("../results/figures/other_plots/KM_plots_signif_cox_all_indiv.pdf", height = 20, width = 15)
grid.arrange(p1, p2, p3, p4, p5, p6, p7, ncol = 2)
#dev.off()
```
\newpage
With log rank test, we see that only Th2 and eosinophils are significant.

### d- Survival analysis for Th2 cells
In this part, we want to see the HR of the genes expressed by Th2 cells To determine if some genes have a dominant effect over others.
```{r, fig.cap="Forest plot for Th2 cells (n = 82)"}
# List of genes expressed in th2 cells
l_genes_th2 <- c("ADCY1", "AHI1", "ANK1", "BIRC5", "CDC25C", "CDC7", "CENPF", "CXCR6", "DHFR", "EVI5", "GATA3", "GSTA4", "HELLS", "IL26", "LAIR2", "LIMA1", "MB", "MICAL2", "NEIL3", "PHEX", "PMCH", "PTGIS", "SLC39A14", "SMAD2", "SNRPD1", "WDHD1")

# Select rows in the expression matrix
df_expr_th2 <- df_expr[rownames(df_expr) %in% l_genes_th2,]
df_expr_th2 <- df_expr_th2[,-c(1:3)]

# Apply coxph model
df_expr_th2 <- as.data.frame(t(df_expr_th2))
df_expr_th2 <- df_expr_th2[df_metadata_surv$Patient,]
df_expr_th2$Patient <- rownames(df_expr_th2)
df_survival_th2 <- merge(df_metadata_surv, df_expr_th2, by = "Patient")
rownames(df_survival_th2) <- df_survival_th2$Patient
df_survival_th2 <- df_survival_th2[, -1]
results_th2 <- apply_coxph_model(df_survival_th2)

# Generate forest plot
#pdf("../results/figures/forest_plots/all_indiv/forest_plot_all_patients_th2.pdf")
generate_forestplot(results_th2, Type = FALSE)
#dev.off()
```
\newpage
So, we see that for th 12 significant genes, 10 have a ratio > 2 et only 2 genes have a small ratio. We now can confirm than the prese,ce of Th2 cells have a bad impact on the prognosis.

### e- Significative immune cells genes survival analysis
Here, we repeat the previous step for the genes of all the significant immune cells.
```{r, fig.height=15, fig.cap="Forest plot for significative immune cells (n = 82)"}
# List of immune cells type significant in coxph analysis
l <- c("DC",  "Eosinophils", "Immune checkpoints", "TFH", "Th2 cells", "Treg", "aDC")

# Select the genes for cell types selected
df_selected_imm <- df_CTA_immune_whole_clean_avg[df_CTA_immune_whole_clean_avg$Signature %in% l, ] %>% select(Signature)
df_selected_imm$Variable <- rownames(df_selected_imm)
df_selected_imm_expr <- df_CTA_immune_whole_clean_avg[, colnames(df_CTA_immune_whole_clean_avg) %in% df_metadata_surv$Patient]
df_selected_imm_expr <- df_selected_imm_expr[rownames(df_selected_imm_expr) %in% rownames(df_selected_imm),]
df_selected_imm_expr <- as.data.frame(t(df_selected_imm_expr))
df_selected_imm_expr$Patient <- rownames(df_selected_imm_expr)

# Merge with metadata
df_selected_imm_expr <- merge(df_metadata_surv[, c("Patient", "OS.delay", "OS.event")], df_selected_imm_expr, by = "Patient")
rownames(df_selected_imm_expr) <- df_selected_imm_expr$Patient
df_selected_imm_expr <- df_selected_imm_expr[, -1]

# Apply coxph model and merge with cel types
results_selected_imm <- apply_coxph_model(df_selected_imm_expr)
results_selected_imm <- merge(results_selected_imm, df_selected_imm, by = "Variable")

# Create forest plot
#pdf("../results/figures/forest_plots/all_indiv/forest_plot_selected_imm_all_indiv.pdf", height = 15, width = 8)
generate_forestplot(results_selected_imm, Type = TRUE)
#dev.off()
```
\newpage
This forest plot show the impact of each cell types on the survival probabilities. We can see the genes that are drive the results.

## 2) Conventional chondrosarcomas
This section concern the survival analysis for conventional chondrosarcomas. This is the same stpes than previously.

### a- Categorical data
```{r, fig.cap="Forest plot for categorical immune cell types expression (n = 63)"}
# Compute medians for each immune cell types
medians_sign <- apply(df_avg_immune_sign_z_scores_conv, 2, median)

# Adding categorical column
df_cat_conv <- df_avg_immune_sign_z_scores_conv
for (col in colnames(df_cat_conv)) {
  new_column_name <- paste(col, "cat", sep = " ")
  df_cat_conv[[new_column_name]] <- ifelse(df_avg_immune_sign_z_scores_conv[[col]] > medians_sign[col], "HIGH", "LOW")
}

# Select categories
df_cat_conv <- df_cat_conv[, grep("cat", colnames(df_cat_conv))]

# Replace space by _
colnames(df_cat_conv) <- gsub(" ", "_", colnames(df_cat_conv))

# Merge df
df_cat_conv$Patient <- rownames(df_cat_conv)
df_survival_conv_cat <- merge(df_metadata_surv_conv, df_cat_conv, by = "Patient")
rownames(df_survival_conv_cat) <- df_survival_conv_cat$Patient
df_survival_conv_cat <- df_survival_conv_cat[, -1]

# Apply coxph model and create forest plot
results <- apply_coxph_model(df_survival_conv_cat)
#write.table(results, "../results/results_coxhp.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
generate_forestplot(results, Type = FALSE)
```
\newpage
By using categorical data and Cox model, there is only Th2 that are signnificant.

### b- Continuous data
```{r, fig.cap="Forest plot for continuous immune cell types expression (n = 63)"}
df_cont_conv <- df_avg_immune_sign_z_scores_conv

# Replace space by _
colnames(df_cont_conv) <- gsub(" ", "_", colnames(df_cont_conv))

# Merge df
df_cont_conv$Patient <- rownames(df_cont_conv)
df_survival_conv <- merge(df_metadata_surv_conv, df_cont_conv, by = "Patient")
rownames(df_survival_conv) <- df_survival_conv$Patient
df_survival_conv <- df_survival_conv[, -1]

# Apply coxph model and generate plot
results_cont <- apply_coxph_model(df_survival_conv)
#write.table(results_cont, "../results/results_coxph_var_cont_patients_conv.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
#pdf("../results/figures/forest_plots/conv_chondro/forest_plot_conv_patients_var_cont.pdf")
generate_forestplot(results_cont, Type = FALSE)
#dev.off()
```
\newpage
By using z-scores, pDC are now significant and have a good impact on survival probabilities.

### c- Kaplan-Meier plots
This shows the graph for each significant HR ratio more than significant to all patients to observe the differences.
```{r, fig.height=20, fig.width=15, fig.cap="Kaplan-Meier plots for immune cell types (n = 63)"}
# Create Kaplan-Meier plots with categorical data
p1 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ DC_cat, data = df_survival_conv_cat), df_survival_conv_cat)
p2 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Eosinophils_cat, data = df_survival_conv_cat), df_survival_conv_cat)
p3 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Immune_checkpoints_cat, data = df_survival_conv_cat), df_survival_conv_cat)
p4 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ TFH_cat, data = df_survival_conv_cat), df_survival_conv_cat)
p5 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Th2_cells_cat, data = df_survival_conv_cat), df_survival_conv_cat)
p6 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Treg_cat, data = df_survival_conv_cat), df_survival_conv_cat)
p7 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ aDC_cat, data = df_survival_conv_cat), df_survival_conv_cat)
p8 <- plot_km(survfit(Surv(OS.delay, OS.event) ~ pDC_cat, data = df_survival_conv_cat), df_survival_conv_cat)
#pdf("../results/figures/other_plots/KM_plots_signif_cox_conv_indiv.pdf", height = 20, width = 15)
grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, ncol = 2)
#dev.off()
```
\newpage
With this test, we see that only Th2 cells are significant. So we see that this type may be play an important role in the survival.

### d- Survival analysis for Th2 cells
```{r, fig.cap="Forest plot for Th2 cells (n = 63)"}
# Take conventional tumors
df_survival_th2 <- merge(df_metadata_surv_conv, df_expr_th2, by = "Patient")
rownames(df_survival_th2) <- df_survival_th2$Patient
df_survival_th2 <- df_survival_th2[, -1]

# Apply coxph model and create forest plot
results_th2 <- apply_coxph_model(df_survival_th2)
#pdf("../results/figures/forest_plots/conv_chondro/forest_plot_conv_th2.pdf")
generate_forestplot(results_th2, Type = FALSE)
#dev.off()
```

### e- Significative immune cells genes survival analysis
```{r, fig.height=6, fig.cap="Forest plot for significative immune cells (n = 63)"}
# List of immune cell types significant
l_th2_pdc <- c("Th2 cells", "pDC")
  
# Select and prepare data
df_selected_imm <- df_CTA_immune_whole_clean_avg[df_CTA_immune_whole_clean_avg$Signature %in% l_th2_pdc, ] %>% select(Signature)
df_selected_imm$Variable <- rownames(df_selected_imm)
df_selected_imm_expr_conv <- df_CTA_immune_whole_clean_avg[, colnames(df_CTA_immune_whole_clean_avg) %in% df_metadata_surv_conv$Patient]
df_selected_imm_expr_conv <- df_selected_imm_expr_conv[rownames(df_selected_imm_expr_conv) %in% rownames(df_selected_imm),]
df_selected_imm_expr_conv <- as.data.frame(t(df_selected_imm_expr_conv))
df_selected_imm_expr_conv$Patient <- rownames(df_selected_imm_expr_conv)
df_selected_imm_expr_conv <- merge(df_metadata_surv[, c("Patient", "OS.delay", "OS.event")], df_selected_imm_expr_conv, by = "Patient")
rownames(df_selected_imm_expr_conv) <- df_selected_imm_expr_conv$Patient
df_selected_imm_expr_conv <- df_selected_imm_expr_conv[, -1]

# Apply coxph model and create forest plot
results_selected_imm_conv <- apply_coxph_model(df_selected_imm_expr_conv)
results_selected_imm_conv <- merge(results_selected_imm_conv, df_selected_imm, by = "Variable")
results_selected_imm_conv <- results_selected_imm_conv[order(results_selected_imm_conv$Signature, results_selected_imm_conv$Pvalue),]
#pdf("../results/figures/forest_plots/conv_chondro/forest_plot_selected_imm_conv_indiv.pdf", height = 8, width = 8)
generate_forestplot(results_selected_imm_conv, Type = TRUE)
#dev.off()
```

\newpage

# II. CTA survival analysis
This section analyzes the impact of CTA on survival probabilites.

## 1) All chondrosarcoma types
This part concern all the patients.
```{r, fig.height=18, fig.cap="Forest plot for significant CTA (n = 82)"}
res_cta <- apply_coxph_model(df_expr_cta)
#write.table(results_cont, "../results/results_coxph_var_cont_cta_zscore.tsv", sep = "\t", row.names = FALSE, quote = FALSE)

# Forest plot on significative CTA
res_cta_signif <- res_cta[res_cta$Pvalue < 0.05,]
#pdf("../results/figures/forest_plots/forest_plot_cta_signif.pdf", width = 10, height = 15)
generate_forestplot(res_cta_signif, Type = FALSE)
#dev.off()
```
\newpage
This forest plot illustrates only the significant HR. We see here that there is more bad impact CTA than good impact. We uses this CTA list in script 5 to analyze expression and make a clustering. This results illustrates that there is groups so we make survival analysis to see the difference between group (see III.)

### a- Intensities jitter plot for significant CTA
In this section, we want to see the intensities per significant CTA to visualize.
```{r, fig.width=15, fig.cap="Intensitites jitter plot for significant CTA (n = 82)"}
# Read intensities
df <- df_CTA_immune_whole_clean_avg[, df_metadata_surv$Patient]
df <- df[rownames(df) %in% res_cta_signif[,1],]
df$SYMBOL <- rownames(df)

# Data transformation for the plot
df_long <- pivot_longer(df, 
                        cols = -SYMBOL, 
                        values_to = "value")

#pdf(file = "../results/figures/other_plots/plot_cta_coxph_signif_all_indiv.pdf", height = 10, width = 50)
ggplot(df_long, aes(x = reorder(SYMBOL, -value), y = value)) +
  geom_jitter(width = 0.2, height = 0, size = 1, alpha = 0.3, colour = "blue") +
  geom_boxplot(alpha = 0.4, outliers = FALSE, colour = "black", fill = NA) +
  labs(title = "Plot of instensities per significant coxph CTA", x = "CTA", y = "Intensities") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
#dev.off()
```



## 2) Conventional chondrosarcomas
This section like part I concerns conventional chondrosarcomas
```{r, fig.height=15, fig.cap="Forest plot for significant CTA (n = 63)"}
# CTA data
df_expr_cta <- df_expr_cta[df_metadata_surv_conv$Patient, ]

# Apply coxph and generate forest plot with significant data
res_cta <- apply_coxph_model(df_expr_cta)
#write.table(results_cont, "../results/results_coxph_var_cont_cta_zscore_all_patietns.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
res_cta_signif <- res_cta[res_cta$Pvalue < 0.05,]
#pdf("../results/figures/forest_plots/conv_chondro/forest_plot_conv_patients_var_cont_cta_zscores_sorted.pdf", width = 10, height = 30)
generate_forestplot(res_cta_signif, Type = FALSE)
#dev.off()
```
\newpage
Same than previously, we see that there is more bad impact CTA than good. With this, we have also clustering these CTA to see groups and make another analysis.

### Intensities jitter plot for significant CTA
```{r, fig.width=15, fig.cap="Intensitites jitter plot for significant CTA (n = 63)"}
# Prepare data
df <- df_CTA_immune_whole_clean_avg[, df_metadata_surv_conv$Patient]
df <- df[rownames(df) %in% res_cta_signif[,1],]
df$SYMBOL <- rownames(df)

# Transform data for the plot
df_long <- pivot_longer(df, 
                        cols = -SYMBOL, 
                        values_to = "value")

# Create the plot
#pdf(file = "../results/figures/other_plots/plot_cta_coxph_signif_conv_indiv.pdf", height = 10, width = 50)
ggplot(df_long, aes(x = reorder(SYMBOL, -value), y = value)) +
  geom_jitter(width = 0.2, height = 0, size = 1, alpha = 0.3, colour = "blue") +
  geom_boxplot(alpha = 0.4, outliers = FALSE, colour = "black", fill = NA) +
  labs(title = "Plot of instensities per significant coxph CTA", x = "CTA", y = "Intensities") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
#dev.off()
```

\newpage

# III. Survival analysis for clusters from CTA heatmaps
This section use clustering results from script 5 on expression analysis.

## 1) All chondrosarcoma types
This part compute with all the patients.

### a- Clustering Kmeans k = 4 with all chondrosarcoma types
The clustering corresponds to the figure 32.
```{r, fig.cap="Kaplan-Meier plot with clusters from figure 32 (script 5) (n = 82)"}
# Read clusters
l_anno_all <- read.table("../results/clusters_indiv/clusters_all_indiv_mhc.tsv", header = TRUE, sep = "\t")
df_cluster_all <- merge(l_anno_all, df_metadata_surv, by = "Patient")
rownames(df_cluster_all) <- df_cluster_all$Patient
p <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Cluster, data = df_cluster_all), df_cluster_all)
p
```
We see that there is no significant differences between the 4 clusters.

### b- Significative coxph CTA genes clustering
This clustering is from heatmap 2.
```{r, fig.cap="Kaplan-Meier plot with clusters from figure 2 (script 5) (n = 82)"}
# Read clusters
l_cta_all <- read.table("../results/clusters_indiv/clusters_all_indiv_mhc_cta_signif_coxph.tsv", header = TRUE, sep = "\t")

df_cluster_cta_all <- merge(l_cta_all, df_metadata_surv, by = "Patient")
rownames(df_cluster_cta_all) <- df_cluster_cta_all$Patient
p <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Cluster, data = df_cluster_cta_all), df_cluster_cta_all)
p
```
Here we see that the C3 have bad survival and this cluster correspond to an over expression of some CTA. So, we can sat that these CTA could be markers.

## 2) Conventional chondrosarcomas
This part concerns conventional types.

### a- Clustering Kmeans k = 2
This clustering is from figure 33.
```{r, fig.cap="Kaplan-Meier plot with clusters from figure 33 (script 5) (n = 63)"}
# Read clusters
l_anno_conv <- read.table("../results/clusters_indiv/clusters_conv_indiv_mhc.tsv", header = TRUE, sep = "\t")
df_cluster_conv <- merge(l_anno_conv, df_metadata_surv_conv, by = "Patient")
rownames(df_cluster_conv) <- df_cluster_conv$Patient
p <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Cluster, data = df_cluster_conv), df_cluster_conv)
p
```
We see that there is not differences.

### b- Significative coxph CTA genes clustering
This clustering is from heatmap 3.
```{r, fig.cap="Kaplan-Meier plot with clusters from figure 3 (script 5) (n = 63)"}
# Read clusters
l_cta_conv <- read.table("../results/clusters_indiv/clusters_conv_indiv_mhc_cta_signif_coxph.tsv", header = TRUE, sep = "\t")

# Prepare table
df_cluster_cta_all <- merge(l_cta_conv, df_metadata_surv, by = "Patient")
rownames(df_cluster_cta_all) <- df_cluster_cta_all$Patient
p <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Cluster, data = df_cluster_cta_all), df_cluster_cta_all)
p
```
This confirm the KM plot with all patients. We clearly see that some CTA have bad impact on survival.

#### Merge C1 and C2
Here, we want to merge the 2 clusters to see the differences than with 3.
```{r, fig.cap="Kaplan-Meier plot with merged clusters from figure 18 (n = 63)"}
# Prepare table
df_cluster_cta_all_merge <- merge(l_cta_conv, df_metadata_surv, by = "Patient")
df_cluster_cta_all_merge$Cluster <- ifelse(df_cluster_cta_all_merge$Cluster == 2, 1, df_cluster_cta_all_merge$Cluster)
rownames(df_cluster_cta_all_merge) <- df_cluster_cta_all_merge$Patient
p <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Cluster, data = df_cluster_cta_all_merge), df_cluster_cta_all_merge)
p
```
We see that there is always significant.

## 3) Cluster 1 from CTA heatmaps analysis
This section is the analysis of the previous results. We want to see where the patients clustered in other heatmaps.
```{r}
# Take indiv from cluster 1
l_c1_cta_all <- l_cta_all[l_cta_all$Cluster == 1, ]

# Merge to see where the indiv are in the other heatmap
df_analysis_cold_all <- l_c1_cta_all %>%
  left_join(df_metadata %>% select(Patient, Histology), by = "Patient") %>%
  left_join(l_anno_all, by = "Patient") 
df_analysis_cold_all
```
We see that there is a lot of dediff type that expressed CTA involved in bad survival events but retrieve in cluster 4 (HOT) for the immune cells. One dediff is in the cold cluster. So the differentiated cancers are really different than conventional and are infiltrated by immune cells or not but the most is infiltrated.

```{r}
# Take indiv from cluster 1
l_c1_cta_conv <- l_cta_conv[l_cta_conv$Cluster == 1, ]
df_analysis_cold_conv <- l_c1_cta_conv %>%
  left_join(df_metadata %>% select(Patient, Histology), by = "Patient") %>%
  left_join(l_anno_all, by = "Patient") %>%
  left_join(l_anno_conv, by = "Patient")
df_analysis_cold_conv
```
For the conventional chondrosarcomas, we see that all the patient have G3 chondrosarcoma and can be in the cluster 2.



