---
title: "Survival analysis of patients with osteosarcoma"
author: "Léa ROGUE"
output:
  pdf_document:
    fig_caption: true
    latex_engine: xelatex
    toc: true
    toc_depth: '4'
    includes:
      before_body: ../../tex_files/list_of_figures.tex
date: "2025-03-13"
---

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(
  tidy.opts = list(width.cutoff = 60), 
  tidy = TRUE, 
  fig.align = 'center'
)
```

\newpage

This script performs survival analysis of osteosarcoma samples (microarray, GSE21257).

# Load librairies
```{r, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(survival)
library(forestplot)
library(survminer)
library(ggplot2)
library(ggsurvfit)
library(gridExtra)
library(tidyr)
library(stringr)
```

# Functions
```{r}
# Function to apply coxph model per columns
apply_coxph_model <- function(df) {
  # Empty df
  results <- data.frame()

  # Loop on each columns to apply coxph
  for (col in colnames(df)[3:ncol(df)]) {
    model <- coxph(Surv(OS.delay, OS.event) ~ df[[col]], data = df)
    
    # Extract results
    exp_coef <- summary(model)$coefficients[,"exp(coef)"]
    lower_ci <- summary(model)$conf.int[,"lower .95"]
    upper_ci <- summary(model)$conf.int[,"upper .95"]
    p_value <- summary(model)$coefficients[,"Pr(>|z|)"]
    results <- rbind(results, data.frame(
      Variable = col,
      HR = exp_coef,
      LowerCI = lower_ci,
      UpperCI = upper_ci,
      Pvalue = p_value
    ))
  }
  return(results)
}

# Function to create Kaplan-Meier plot
plot_km <- function(mod, df) {
  km_plot <- ggsurvplot(mod,
             data = df,      
             pval = TRUE,
             conf.int = FALSE,
             ggtheme = theme_bw(),
             palette = c("#E7B800", "#2E9FDF", "#FF6F61", "#4EBB92"))
    return(km_plot$plot)
}

# Function to generate forest plot from coxph analysis from the function apply_coxph_model
generate_forestplot <- function(data, Type) {
  if (Type == TRUE){
    # Order by Signature and p-value
    data <- data[order(data$Signature,  data$Pvalue),]
    data$Pvalue <- sprintf("%.4f", data$Pvalue)
    data %>%
      forestplot(
        labeltext = c(Signature, Variable, Pvalue),   
        mean = HR,
        lower = LowerCI,
        upper = UpperCI,
        grid = TRUE,
        zero = 1,
        col = fpColors(box = "black", line = "black"),
        hrzl_lines = TRUE,
        title = "Forest Plot for Cox Model",
        txt_gp = fpTxtGp(label = gpar(fontsize = 8)),
        ci.vertices = TRUE,
        boxsize = 0.1) %>%
      fp_set_zebra_style("#EFEFEF") %>% 
      fp_add_header(
        Signature = c("", "Cell type"),
        Variable = c("", "Genes"),
        Pvalue = c("", "p-value"))
  }
  else {
    # Order by p-value
    data <- data[order(data$Pvalue),]
    data$Pvalue <- sprintf("%.4f", data$Pvalue)
    data %>%
      forestplot(
        labeltext = c(Variable, Pvalue),   
        mean = HR,
        lower = LowerCI,
        upper = UpperCI,
        grid = TRUE,
        zero = 1,
        col = fpColors(box = "black", line = "black"),
        hrzl_lines = TRUE,
        title = "Forest Plot for Cox Model",
        txt_gp = fpTxtGp(label = gpar(fontsize = 8)),
        ci.vertices = TRUE,
        boxsize = 0.1) %>%
      fp_set_zebra_style("#EFEFEF") %>% 
      fp_add_header(
        Variable = c("", "Genes"),
        Pvalue = c("", "p-value"))    
  }
}
```

# Load and format data
```{r}
# Matrix with z-scores intensities and information on immune signature and CTA
df_expr_z_scores <- read.table("../results/expr_matrix_CTA_sign_imm_z_scores.tsv", sep = "\t", header = TRUE, check.names = FALSE)

# Matrix with average expression per cell types in z-scores
df_imm_z_scores <- read.table("../results/imm_sign_avg_z_scores.tsv", sep = "\t", header = TRUE, check.names = FALSE)
rownames(df_imm_z_scores) <- df_imm_z_scores$Signature
df_imm_z_scores <- as.data.frame(t(df_imm_z_scores[, -1]))
```


```{r}
# Create a better matrix
metadata <- read.table("../data/metadata.tsv", sep = "\t", header = F)
df_metadata <- as.data.frame(t(metadata))
colnames(df_metadata) <- df_metadata[1,]
df_metadata <- df_metadata[-1,]
colnames(df_metadata) <- gsub(" ", "_", colnames(df_metadata))
df_metadata$OS.event <- ifelse(grepl("Alive", df_metadata$status), 0, 1)
df_metadata$OS.delay <- as.numeric(str_extract(df_metadata$status, "\\d+"))
df_metadata_surv <- df_metadata %>% select(Patient, OS.delay, OS.event)
```

\newpage

# I. Immune cells survival analysis
This part concern the survival analysis with the expression of the immune cells. This is to observe the impact of each cell types on the survival probabilities of the patients.

## a- CoxPH model
In this section, we use z-scores data with Cox model.
```{r, fig.cap="Forest plot for continuous immune cell types expression (n = 53)"}
# Continuous data
df_cont_all <- df_imm_z_scores

# Replace space by _
colnames(df_cont_all) <- gsub(" ", "_", colnames(df_cont_all))

# Merge df
df_cont_all$Patient <- rownames(df_cont_all)
df_survival_cont <- merge(df_metadata_surv, df_cont_all, by = "Patient")
rownames(df_survival_cont) <- df_survival_cont$Patient
df_survival_cont <- df_survival_cont[, -1]


# Apply coxph on continuous
results_cont <- apply_coxph_model(df_survival_cont)
#write.table(results_cont, "../results/results_coxph_var_cont.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
#pdf("../results/figures/forest_plots/all_indiv/forest_plot_all_patients_var_cont.pdf")
generate_forestplot(results_cont, Type = FALSE)
#dev.off()
```

\newpage

# II. CTA survival analysis
This section analyzes the impact of CTA on survival probabilites.
```{r, fig.height=10, fig.cap="Forest plot for significant CTA"}

df_expr_cta <- subset(df_expr_z_scores, CTA == "CTA")
rownames(df_expr_cta) <- df_expr_cta$SYMBOL
df_expr_cta <- df_expr_cta[, -c(1:3)]
df_expr_cta <- as.data.frame(t(df_expr_cta))
df_expr_cta$Patient <- rownames(df_expr_cta)
df_expr_cta <- merge(df_metadata_surv, df_expr_cta, by = "Patient")
rownames(df_expr_cta) <- df_expr_cta$Patient
df_expr_cta <- df_expr_cta[, -1]

res_cta <- apply_coxph_model(df_expr_cta)

# Forest plot on significative CTA
res_cta_signif <- res_cta[res_cta$Pvalue < 0.05,]
#write.table(res_cta_signif, "../results/results_coxph_osteo_cta_zscore_signif.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
#pdf("../results/figures/forest_plots/forest_plot_cta_signif.pdf", width = 10, height = 15)
generate_forestplot(res_cta_signif, Type = FALSE)
#dev.off()
```
\newpage
This forest plot illustrates only the significant HR. We see here that there is more bad impact CTA than good impact. 

\newpage


# III. Survival analysis for clusters from CTA heatmaps
This section use clustering results from script 1 on expression analysis.

## 1) Significative coxph CTA genes clustering
This clustering is from heatmap 3.
```{r, fig.cap="Kaplan-Meier plot with 4 clusters"}
# Read clusters
l_clust <- read.table("../results/clusters_cta_signif_coxph_osteo_53.tsv", header = TRUE, sep = "\t")

# Prepare table
df_cluster_cta_all <- merge(l_clust, df_metadata_surv, by = "Patient")
rownames(df_cluster_cta_all) <- df_cluster_cta_all$Patient
p <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Cluster, data = df_cluster_cta_all), df_cluster_cta_all)
p
```
This confirm the KM plot with all patients. We clearly see that some CTA have bad impact on survival.

#### Merge C2 and C3
Here, we want to merge the 2 clusters to see the differences than with 3.
```{r, fig.cap="Kaplan-Meier plot with 3 clusters (2+3)"}
# Prepare table
df_cluster_cta_all_merge <- merge(l_clust, df_metadata_surv, by = "Patient")
df_cluster_cta_all_merge$Cluster <- ifelse(df_cluster_cta_all_merge$Cluster == 2, "2+3", df_cluster_cta_all_merge$Cluster)
df_cluster_cta_all_merge$Cluster <- ifelse(df_cluster_cta_all_merge$Cluster == 3, "2+3", df_cluster_cta_all_merge$Cluster)
rownames(df_cluster_cta_all_merge) <- df_cluster_cta_all_merge$Patient
p <- plot_km(survfit(Surv(OS.delay, OS.event) ~ Cluster, data = df_cluster_cta_all_merge), df_cluster_cta_all_merge)
p
```
We see that it is significant.



